---
title: "Tools to Interact with the MS Bioscreen API"
author: ""
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tools to Interact with the MS Bioscreen API}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Setting an authorization token

Before interacting with the MS Bioscreen API, you must have a valid authorization token to access the various API urls.

* Almost all msbwaiter functions rely on an authorization token. In these functions, the parameter `token` is set to `get_token()` as default. 

* `get_token` conveniently allows you to set a token once at the beginning of an R session. Running `get_token()` returns the value of the environment variable `MSBWAITER_TOKEN`: 

```{r, eval = FALSE}

Sys.setenv("MSBWAITER_TOKEN" = "MY_TOKEN")

get_token()
# [1] "MY_TOKEN"

```

* If the environment variable `MSBWAITER_TOKEN` is not set, `get_token` will prompt you to enter a token:

```{r, eval = FALSE}

Sys.unsetenv("MSBWAITER_TOKEN")

get_token()
# Please enter your authorization token (without quotes): MY_TOKEN

```

* If you enter an empty token, `get_token` will continue to prompt you to enter a token:

```{r, eval = FALSE}

Sys.unsetenv("MSBWAITER_TOKEN")

get_token()
# Please enter your authorization token (without quotes):
# Sorry, this is an empty token...
# Please enter your authorization token (without quotes):
# Sorry, this is an empty token...
# Please enter your authorization token (without quotes): MY_TOKEN

```

* You can change the value of the environment variable `MSBWAITER_TOKEN` with `overwrite = TRUE`:

```{r, eval = FALSE}

Sys.setenv("MSBWAITER_TOKEN" = "MY_TOKEN")

get_token(overwrite = TRUE)
# Current token is:
# MY_TOKEN
# Please enter your new authorization token (without quotes), or press enter to keep existing one: NEW_TOKEN

get_token()
# [1] "NEW_TOKEN"

```

## Querying

msbwaiter provides a number functions to query specific sets of data from the bioscreen

* `api_paginate_action` is a general function that can be used to search for different sets of data

* When `action = GET`, `api_paginate_action` performs a batch query, returning all the data associated with the specified `endpoint`. Note that the parameter `search_by_list` is ignored when `action = GET`.

```{r, eval = FALSE}

# return all the subjects data currently in the bioscreen
api_paginate_action(action = GET, endpoint = "subjects")

# return all the attacks data currently in the bioscreen
api_paginate_action(action = GET, endpoint = "attacks")

```

* When `action = POST`, `api_paginate_action` searches for the data that matches the specified `search_by_list`, a list of search parameter names and values.

```{r, eval = FALSE}

# to search for the subject whose source_id = 1 and external_identifier = 1...
search_by_list = list(source_id = 1, external_identifier = 1)
api_paginate_action(action = POST, search_by_list = search_by_list, endpoint = "subjects")
  
```

* Use the parameters `number_of_pages` and `results_per_page` to limit the amount of data that is returned

```{r, eval = FALSE}

# get only the first 2 pages of attacks data...
api_paginate_action(action = GET, endpoint = "attacks", number_of_pages = 2)

# speed up how long it takes to fetch each page by setting results_per_page to 100...
api_paginate_action(action = GET, endpoint = "attacks", results_per_page = 100)

# get the first 200 entries of attacks data in multiple ways...
api_paginate_action(action = GET, endpoint = "attacks", number_of_pages = 2, results_per_page = 100)
api_paginate_action(action = GET, endpoint = "attacks", number_of_pages = 1, results_per_page = 200)

```

* `api_get_batch` performs a batch query, returning all the data associated with the specified endpoint. It essentially runs `api_paginate_action` with `action = GET`.

* As in `api_paginate_action`, you can set `number_of_pages` and `results_per_page` to limit the amount of data that is returned.

```{r, eval = FALSE}

# get all the subjects data
api_get_batch(endpoint = "subjects")

# get the first 3 pages of treatments data
api_get_batch(endpoint = "treatments", number_of_pages = 3)

# get the first 800 entries of visits data
api_get_batch(endpoint = "visits", number_of_pages = 4, results_per_page = 200)

```

* `api_get` returns the data associated with the specified `source_id`, `external_identifier`, and `endpoint`. `api_get` will always return 1 row of data.

```{r, eval = FALSE}

# get the attack where source_id = 1 and external_identifier = 1
api_get(source_id = 1, external_identifier = 1, endpoint = "attacks")

```

* `api_search_by_msbid` returns the entries associated with the specified `msbid` and `endpoint`.

```{r, eval = FALSE}

# get all the treatments for the subject whose msbid = 28
api_search_by_msbid(msbid = 28, endpoint = "treatments")

```

* `api_search_by_updated_at` returns the entries updated on or after the specified `updated_at_date` and `updated_at_time`.

```{r, eval = FALSE}

# get all the visits updated on or after 2015-12-01 at midnight
api_search_by_updated_at(updated_at_date = "2015-12-01", updated_at_time = "00:00:00", 
                         endpoint = "visits")

```

## Creating

You can use `api_create` to easily create a new entry in the bioscreen.

* `api_create` takes `sufl_data` as an input. `sufl_data` is a data frame with 1 row and column names that follow the current [SUFL specification](https://docs.google.com/spreadsheets/d/1SBe_p6EcdPeus1WTvN3UbIvJC1PtI2ntYowSwfAdsEQ/edit?pref=2&pli=1#gid=248464688). At minimum, `sufl_data` must contain the identifer columns **source_id** and **external_identifier**. For attacks, treatments and visits data, `sufl_data` must also contain the identifier columns **patient_source_id** and **patient_external_identifier**.

```{r, eval = FALSE}

# creating a subject
sufl_data = data.frame(source_id = 1, external_identifer = 1, gender = "F", 
                       age_of_onset = 33, relapse_count_onset_to_year_5 = 5)
api_create(sufl_data = sufl_data, endpoint = "subjects")

# creating an attack
sufl_data = data.frame(source_id = 1, external_identifer = 1, 
                       patient_source_id = 1, patient_external_identifier = 1,
                       year = 1999, month = 8, day = 1, optic_neuritis = TRUE)
api_create(sufl_data = sufl_data, endpoint = "attacks")

# creating a treatment
sufl_data = data.frame(source_id = 1, external_identifer = 1, 
                       patient_source_id = 1, patient_external_identifier = 1, 
                       start_date = "2011-02-01", name = "Gilenya")
api_create(sufl_data = sufl_data, endpoint = "treatments")

# creating a visit
sufl_data = data.frame(source_id = 1, external_identifer = 1, 
                       patient_source_id = 1, patient_external_identifier = 1, 
                       date = "2012-02-22", disease_course = "RR", edss = 3)
api_create(sufl_data = sufl_data, endpoint = "visits")

```

## Updating

msbwaiter provides `api_update` to update an existing bioscreen entry's values.

* `api_update` uses the identifier columns in `sufl_data` to find the corresponding entry in the bioscreen and updates that entry with the values in `sufl_data`.

```{r, eval = FALSE}

sufl_data = data.frame(source_id = 1, external_identifer = 1, gender = "F", 
                       age_of_onset = 30, relapse_count_onset_to_year_5 = 3)
api_update(sufl_data = sufl_data, endpoint = "subjects")

```

* However, if you want to ensure that certain variables do not get updated in the bioscreen even if their values differ from those in `sufl_data`, use `ignore_colnames` to specify those variables. Specifying these variables with `ignore_colnames` is an alternative to removing these columns entirely from `sufl_data`.

```{r, eval = FALSE}

sufl_data = data.frame(source_id = 1, external_identifer = 1, gender = "F", 
                       age_of_onset = 30, relapse_count_onset_to_year_5 = 3)

# with ignore_colnames set to c("gender", "relapse_count_onset_to_year_5"), api_update will not 
# update the values for gender and relapse_count_onset_to_year_5 in the bioscreen even if those
# values are not equal to "F" and 3.
api_update(sufl_data = sufl_data, endpoint = "subjects", 
           ignore_colnames = c("gender", "relapse_count_onset_to_year_5"))

```

* You can also specify whether variables with non-missing values in the bioscreen should be overwritten to missing if these variables are equal to NA in `sufl_data`. With `keep_na = TRUE`, non-missing values can be overwritten to missing (the default is set to `FALSE`).

```{r, eval = FALSE}

sufl_data = data.frame(source_id = 1, external_identifer = 1, gender = "F", 
                       age_of_onset = NA, relapse_count_onset_to_year_5 = 3)

# if the value for age_of_onset in the bioscreen should NOT be overwritten to missing even though 
# age_of_onset is equal to NA in sufl_data, set keep_na to FALSE (default).
api_update(sufl_data = sufl_data, endpoint = "subjects", keep_na = FALSE)

# if the value for age_of_onset in the bioscreen should be overwritten to missing even if it is 
# currently non-missing value in the bioscreen, set keep_na to TRUE.
api_update(sufl_data = sufl_data, endpoint = "subjects", keep_na = TRUE)

```

## Deleting

msbwaiter provides `api_delete` and `api_delete_batch` to delete entries in the bioscreen. Use these functions with caution!

* `api_delete` deletes the entry associated with the specified `source_id`, `external_identifier`, and `endpoint`.

* Note that when `endpoint = "subjects"`, `api_delete` deletes the specified subject along with any attacks, treatments, and visits associated with this subject.

```{r, eval = FALSE}

# to delete the subject whose source_id = 1 and external_identifier = 1 AND all 
# attacks, treatments, and visits associated with this subject...
api_delete(source_id = 1, external_identifier = 1, endpoint = "subjects")

# to delete the visit whose source_id = 1 and external_identifier = 1...
api_delete(source_id = 1, external_identifier = 1, endpoint = "visits")

```

* `api_delete_batch` deletes all the entries associated with the specified `source_ids` and `endpoint`.

* As with `api_delete`, when `endpoint = "subjects"`, `api_delete_batch` deletes the specified subjects along with any attacks, treatments, and visits associated with these subjects.

```{r, eval = FALSE}

# to delete all the subjects whose source_id = 1 AND all 
# attacks, treatments, and visits associated with these subjects...
api_delete_batch(source_ids = 1, endpoint = "subjects")

# to delete all the visits whose source_id = 1...
api_delete_batch(source_ids = 1, endpoint = "visits")

```

## Querying before creating, updating, or deleting

`api_check` and `api_check_batch` allow you to interact with the bioscreen API in a smart way. These functions check what data is already present in the bioscreen before proceeding to create, update, or delete entries.

* Use `api_check` for creating or updating a single bioscreen entry. Like `api_create` and `api_update`, `api_check` takes `sufl_data` as an input.

* `api_check` uses the identifier columns of `sufl_data` to find the corresponding entry in the bioscreen. If no corresponding entry exists, `api_check` determines it needs to create that entry. If a corresponding entry exists but the values between `sufl_data` and the corresponding entry do not match, `api_check` determines it needs to update that entry. If the values between `sufl_data` and the corresponding entry do match, `api_check` returns "no action".

* When `change = TRUE`, `api_check` proceeds to actually create or update the entry or return "no action". When `change = FALSE`, `api_check` simply returns a character string that is equal to either "create", "update", or "no action".

```{r, eval = FALSE}

sufl_data = data.frame(source_id = 1, external_identifer = 1, gender = "F", 
                       age_of_onset = 33, relapse_count_onset_to_year_5 = 5)

# checking if the subject with source_id = 1 and external_identifier = 1 needs to 
# be created, updated, or if the most recent data is already present in the bioscreen

# ...without actually creating or updating the entry
api_check(sufl_data = sufl_data, endpoint = "subjects", change = FALSE)

# ...actually creating or updating the entry if necessary
api_check(sufl_data = sufl_data, endpoint = "subjects", change = TRUE)

```

* Like `api_update`, `api_check` has the parameters `ignore_columnames` and `keep_na` available to detail what variables should be taken into consideration when determining whether an entry needs to be updated.

```{r, eval = FALSE}

sufl_data = data.frame(source_id = 1, external_identifer = 1, gender = "F", 
                       age_of_onset = NA, relapse_count_onset_to_year_5 = 3)

# with ignore_colnames set to c("gender", "relapse_count_onset_to_year_5"), api_check will
# ignore the values for gender and relapse_count_onset_to_year_5 when determining whether
# the entry needs to be updated. api_check will not update these values in the bioscreen when
# change = TRUE.
api_check(sufl_data = sufl_data, endpoint = "subjects", 
          ignore_colnames = c("gender", "relapse_count_onset_to_year_5"))

# with keep_na set to FALSE (default), api_check will ignore age_of_onset when determing whether
# the entry needs to be updated. api_check will NOT overwrite age_of_onset to missing in the bioscreen
# when change = TRUE.
api_check(sufl_data = sufl_data, endpoint = "subjects", keep_na = FALSE)

# with keep_na set to TRUE, api_check will consider age_of_onset when determing whether
# the entry needs to be updated. If age_of_onset is non-missing in the bioscreen, api_check
# will determine that the entry needs to be updated. api_check will overwrite age_of_onset to missing
# when change = TRUE.
api_check(sufl_data = sufl_data, endpoint = "subjects", keep_na = TRUE)

```

* `api_check_batch` works similarly to `api_check`. It takes a data frame with appropriate identifier columns and column names that follow the current SUFL specification. However, unlike `api_check`, `api_check_batch` takes a data frame, called `sufl_batch`, with an unlimited number of rows. msbwaiter provides `api_check_batch` to create, update, and delete multiple bioscreen entries at once.

* `api_check_batch` finds the entries that correspond to each of the entries in `sufl_batch` and, similarly to `api_check`, performs a comparison on each entry. For each entry, `api_check_batch` returns either "create", "update", or "no action". When `change = TRUE`, `api_check_batch` proceeds to create and update entries.

* Like `api_check`, `api_check_batch` has the paramters `ignore_columnames` and `keep_na` available to detail what variables should be taken into consideration when determining which entries need to be updated.

```{r, eval = FALSE}

sufl_batch = data.frame(source_id = c(1,1,3), external_identifer = c(1, 2,100), 
                       gender = c("F", "M", "F"), age_of_onset = c(NA, 19, 35), 
                       relapse_count_onset_to_year_5 = c(3, 4, 2))

# check to see if any of the entries in sufl_batch exist in the bioscreen
# create the entries if they do not exist, update the entries if the values
# are different
api_check_batch(sufl_batch, endpoint, change = TRUE)

# with ignore_columnames = c("gender"), api_check_batch will not update
# gender even if the value for gender in the bioscreen is different than
# that in sufl_batch
api_check_batch(sufl_batch, endpoint, ignore_colnames = c("gender"), change = TRUE)

# with keep_na = TRUE, sufl_batch will overwrite any non-missing values 
# in the bioscreen to missing if that value = NA in sufl_batch 
api_check_batch(sufl_batch, endpoint, keep_na = TRUE, change = TRUE)

```

* Unique to `api_check_batch` are the parameters `destructive`, `max_update`, and `max_delete`

* When `change = TRUE` and `destructive = TRUE`, `api_check_batch` will delete any entries in the bioscreen that are not found in `sufl_batch` but that have a source_id that is found in `sufl_batch`.

```{r, eval = FALSE}

sufl_batch = data.frame(source_id = c(1,1,3), external_identifer = c(1, 2,100), 
                       gender = c("F", "M", "F"), age_of_onset = c(NA, 19, 35), 
                       relapse_count_onset_to_year_5 = c(3, 4, 2))

# with destructive = TRUE, api_check_batch will delete any entries in the bioscreen
# that have a source_id that is found in sufl_batch but whose external_identifier 
# is not found in sufl_batch
api_check_batch(sufl_batch, endpoint, change = TRUE, destructive = TRUE)

```

* `max_update` and `max_delete` are safety parameters. If `change = TRUE` but the the number of entries to update exceeds `max_update`, `api_check_batch` will not proceed to create, update, or delete entries. Similarly, if `change = TRUE` and `destructive = TRUE` but the the number of entries to delete exceeds `max_delete`, `api_check_batch` will not proceed to create, update, or delete entries.

```{r, eval = FALSE}

sufl_batch = data.frame(source_id = c(1,1,3), external_identifer = c(1, 2,100), 
                       gender = c("F", "M", "F"), age_of_onset = c(NA, 19, 35), 
                       relapse_count_onset_to_year_5 = c(3, 4, 2))

# with max_update = 2, api_check_batch will not proceed to create, update, or delete
# entries if more than 2 entries in the bioscreen need to be updated.
api_check_batch(sufl_batch, endpoint, change = TRUE, max_update = 2)

# with max_delete = 5, api_check_batch will not proceed to create, update, or delete
# entries if more than 5 entries in the bioscreen need to be updated.
api_check_batch(sufl_batch, endpoint, change = TRUE, destructive = TRUE, max_delete = 5)

```

